---
title: "Investigating the impact of severe weather events on public health and 
the economy"
author: "Ioannis Moros"
date: "June 18, 2017"
output: html_document
---

# Synopsis
Storms and other severe weather events can cause both public health and economic 
problems for communities and municipalities. Many severe events can result in 
fatalities, injuries, and property damage, and preventing such outcomes to the 
extent possible is a key concern.

This paper explores the U.S. National Oceanic and Atmospheric Administration's 
(NOAA) storm database. This database tracks characteristics of major storms and 
weather events in the United States, including when and where they occur, as 
well as estimates of any fatalities, injuries, and property damage. 

The aim of this analysis is to identify which types of events are most harmful 
with respect to population health and which
types of events have the greatest economic consequences.

The results of the analysis show that tornadoes are the most harmful weather event
with respect to population health, in terms of both injuries and fatalities. 
Regarding the total economic impact, floods are the most destructive event.
Most property losses are attributed to floods, while drought is responsible 
for the most damage to crops.

# R/Knit setup

Before the actual data processing we define the knitr parameters and require the
necessary R packages.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

```{r pack}
## Required packages
require(plyr)
require(reshape2)
require(ggplot2)
```

# Data Processing

1. Download and read the raw data.

```{r dl}

if(!file.exists("./data")) {
        dir.create("./data")
        url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        file_dest <- "./data/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(url, file_dest)
}

A <- read.csv("./data/repdata%2Fdata%2FStormData.csv.bz2")
```

2. Subset the raw table to include only the variables of interest

```{r subsetA.1}
Asub <- subset(A, select = c("EVTYPE", "FATALITIES", "INJURIES",
                             "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))
```

3. Investigate the variables that hold the exponential component of economic
and health damage.

```{r summary.1}
summary(Asub$PROPDMGEXP)
summary(Asub$CROPDMGEXP)
```

4. In order to get the actual value of the damage, these 2 
variables will need to be converted to numeric, then multiplied by the respective
value of damage. Notice that the variables contain both uppercase and
lower case values. As such, the first step is to convert all the values to 
uppercase.

```{r uppercase}
Asub$PROPDMGEXP <- toupper(Asub$PROPDMGEXP)
Asub$CROPDMGEXP <- toupper(Asub$CROPDMGEXP)
```

5. Now we are ready to convert their values to numeric.

```{r numeric.transf}
Asub$CROPDMGEXP <- revalue(Asub$CROPDMGEXP, c("?" = 1,
                                        "0" = 1,
                                        "2" = 10^2,
                                        "K" = 10^3,
                                        "M" = 10^6,
                                        "B" = 10^9
))

Asub$CROPDMGEXP <- as.numeric(as.character(Asub$CROPDMGEXP))
Asub$CROPDMGEXP[is.na(Asub$CROPDMGEXP)] <- 1

Asub$PROPDMGEXP <- revalue(Asub$PROPDMGEXP, c("?" = 1,
                                              "+" = 1,
                                              "-" = 1,
                                              "0" = 1,
                                              "1" = 1,
                                              "2" = 10^2,
                                              "H" = 10^2,
                                              "3" = 10^3,
                                              "K" = 10^3,
                                              "4" = 10^4,
                                              "5" = 10^5,
                                              "6" = 10^6,
                                              "M" = 10^6,
                                              "7" = 10^7,
                                              "8" = 10^8,
                                              "B" = 10^9      
))

Asub$PROPDMGEXP <- as.numeric(as.character(Asub$PROPDMGEXP))
Asub$PROPDMGEXP[is.na(Asub$PROPDMGEXP)] <- 1
```
        
6. Let's take a look at the levels of the EXP variables now.

```{r exp.levels}
levels(as.factor(Asub$CROPDMGEXP))
levels(as.factor(Asub$PROPDMGEXP))
```
 
7. Calculate the total amount of damage and store it as a new variable within 
the same matrix.

```{r totaldmg}
Asub$cropdamage <- Asub$CROPDMG * Asub$CROPDMGEXP
Asub$propdamage <- Asub$PROPDMG * Asub$PROPDMGEXP
```

8. First for the economic damage - Create 3 aggregate matrices with the total 
crop/property/total economic damage by type of event. 
Then create a 4th matrix that combines the other 3 (column-bound) 
and rename the columns.

```{r aggr}
Crop_dmg_agr <- aggregate(cropdamage ~ EVTYPE, data = Asub, FUN = sum)
Prop_dmg_agr <- aggregate(propdamage ~ EVTYPE, data = Asub, FUN = sum)
Econ_dmg_agr <- aggregate(cropdamage + propdamage ~ EVTYPE, data = Asub, FUN = sum)

Econ_dmg_full <- cbind(Econ_dmg_agr, Crop_dmg_agr[2], Prop_dmg_agr[2])
colnames(Econ_dmg_full) <- c("event", "total", "crop", "property")
``` 

9. Order the matrices created in step 8 by damage (descending). Then, only keep
the first 10 rows of the full economic damage matrix, i.e. the top 10 most 
impactful types of events with respect to all types of economic damage.

```{r order}
Crop_dmg_agr <- arrange(Crop_dmg_agr, desc(cropdamage))
Prop_dmg_agr <- arrange(Prop_dmg_agr, desc(propdamage))
Econ_dmg_agr <- arrange(Econ_dmg_agr, desc(Econ_dmg_agr[2]))
Econ_dmg_full <- arrange(Econ_dmg_full, desc(Econ_dmg_full[2]))

Econ_dmg_full <- Econ_dmg_full[1:10, ]
```

10. Create a long version of the full combined matrix. This will be useful later 
to construct the plots.

```{r melt}
E_melt <- melt(data = Econ_dmg_full, id.vars = "event", value.name = "damage")
```

11. Repeat steps 8-10 to create and transform similar matrices for health damage.

```{r health.trans}
Injuries_agr <- aggregate(INJURIES ~ EVTYPE, data = Asub, FUN = sum)
Fatalities_agr <- aggregate(FATALITIES ~ EVTYPE, data = Asub, FUN = sum)
Health_dmg_agr <- aggregate(FATALITIES + INJURIES ~ EVTYPE, data = Asub, FUN = sum)

Health_dmg_full <- cbind(Health_dmg_agr, Injuries_agr[2], Fatalities_agr[2])
colnames(Health_dmg_full) <- c("event", "total", "injuries", "fatalities")

Injuries_agr <- arrange(Injuries_agr, desc(INJURIES))
Fatalities_agr <- arrange(Fatalities_agr, desc(FATALITIES))
Health_dmg_agr <- arrange(Health_dmg_agr, desc(Health_dmg_agr[2]))
Health_dmg_full <- arrange(Health_dmg_full, desc(Health_dmg_full[2]))

Health_dmg_full <- Health_dmg_full[1:10, ]

H_melt <- melt(data = Health_dmg_full, id.vars = "event", value.name = "count")
```

# Results

1. Weather events responsible for the most damage to crops.

```{r cropdmg}
head(Crop_dmg_agr, 5)
```

2. Weather events that have caused the most property losses.

```{r propdmg}
head(Prop_dmg_agr, 5)
```

3. Weather events that have caused the most total economic damage

```{r econdmg}
head(Econ_dmg_agr, 5)
```

4. Weather events responsible for the most fatalities

```{r fatal}
head(Fatalities_agr, 5)
```

5. Weather events that have caused the most injuries

```{r injur}
head(Injuries_agr, 5)
```

6. Weather events that have the biggest impact on population health

```{r healthdmg}
head(Health_dmg_agr, 5)
```



This can be visualised better in the following two plots.

7. Plot of economic damage by type of event.

```{r econplot}
g <- ggplot(E_melt, aes(event, damage, fill = variable))
g + geom_bar(position = "dodge", stat = "identity") + 
        theme(legend.title = element_blank(), 
              axis.text.x = element_text(angle = 45, hjust = 1)) +
        ggtitle("Economic damage by type of event")
```

8. Plot of health damage by type of event.

```{r healthplot}
p <- ggplot(H_melt, aes(event, count, fill = variable))
p + geom_bar(position = "dodge", stat = "identity") + 
        theme(legend.title = element_blank(), 
              axis.text.x = element_text(angle = 45, hjust = 1)) +
        ggtitle("Health damage by type of event")
```

The results presented above reveal that tornadoes are the most harmful weather event
with respect to population health, in terms of both injuries and fatalities. 
Regarding the total economic impact, floods are the most destructive weather event
overall. Most properies have been lost due to floods, while most crops are destroyed
by drought.